---
title: "bedbug_plots_2026"
author: "Montse Robles"
date: "2026-02-25"
output: html_document
---

```{r setup, include=FALSE}
setwd("/Users/roblesgi/Documents/roblesgi/bedbug/rnaseq/")
outdir <- "Figures_2026"
  library (tidyverse)
  library(ggplot2)
  library(ggrepel)
  library(UpSetR)
  library(openxlsx)
```

We're creating a new folder for the final plots 2026 so we can then edit them on Afinnity and photoshop

```{r}
load("deseq_objects_for_figures.RData")
```

Settings for the comparisons
```{r}
treatments_fig <- c("Carvacrol", "Deltamethrin", "PBO", "Thymol")
padj_cut <- 0.05
lfc_cut  <- 1
top_n_labels <- 10
```

helpers for themes
```{r}
theme_clean <- function() {
  theme_bw() +
    theme(
      panel.grid = element_blank(),
      plot.title = element_blank(),
      legend.position = "none"
    )
}
```

Volcano plot settings (themes)

```{r}
# VOLCANO HELPERS (pvalue on Y, highlight DEGs with padj)

make_volcano_df <- function(res_obj) {
  as.data.frame(res_obj) %>%
    rownames_to_column("gene_id") %>%
    filter(!is.na(log2FoldChange)) %>%
    mutate(
      pvalue_plot = if_else(is.na(pvalue), 1, pvalue),
      pvalue_plot = pmax(pvalue_plot, 1e-300),  # avoid -Inf
      padj_plot   = if_else(is.na(padj), 1, padj)
    )
}

make_volcano_plot <- function(df, label = FALSE, top_n = 10, padj_cut = 0.05, lfc_cut = 1) {

  df <- df %>%
    mutate(sig = padj_plot < padj_cut & abs(log2FoldChange) >= lfc_cut)

  top_genes <- df %>%
    filter(sig) %>%
    arrange(padj_plot) %>%          # pick labels using padj (not pvalue)
    slice_head(n = top_n)

  p <- ggplot(df, aes(x = log2FoldChange, y = -log10(pvalue_plot))) +
    geom_point(aes(alpha = sig), color = "black", size = 2) +
    scale_alpha_manual(values = c(`TRUE` = 0.9, `FALSE` = 0.25)) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +  # raw p=0.05 reference
    geom_vline(xintercept = c(-lfc_cut, lfc_cut), linetype = "dashed", color = "black") +
    labs(x = "log2 fold change", y = "-log10 p-value") +
    theme_clean()

  if (label && nrow(top_genes) > 0) {
    p <- p + geom_text_repel(
      data = top_genes,
      aes(label = gene_id),
      color = "black",
      size = 4,
      max.overlaps = Inf
    )
  }

  p
}

save_volcano_pair <- function(df, out_stub, top_n = 10, padj_cut = 0.05, lfc_cut = 1) {
  p_lab <- make_volcano_plot(df, label = TRUE,  top_n = top_n, padj_cut = padj_cut, lfc_cut = lfc_cut)
  p_nol <- make_volcano_plot(df, label = FALSE, top_n = top_n, padj_cut = padj_cut, lfc_cut = lfc_cut)

  ggsave(paste0(out_stub, "_labels.png"),   p_lab, width = 10, height = 8)
  ggsave(paste0(out_stub, "_labels.pdf"),   p_lab, width = 10, height = 8)
  ggsave(paste0(out_stub, "_nolabels.png"), p_nol, width = 10, height = 8)
  ggsave(paste0(out_stub, "_nolabels.pdf"), p_nol, width = 10, height = 8)
}
```

UpSet helpers
```{r}
# UpSet HELPERS (original-style, reliable file output + render in HTML)

deg_ids_from_res <- function(res_obj, padj_cut = 0.05, lfc_cut = NULL) {
  df <- as.data.frame(res_obj) %>%
    rownames_to_column("gene_id") %>%
    mutate(padj_use = if_else(is.na(padj), 1, padj))

  if (is.null(lfc_cut)) {
    df %>% filter(padj_use < padj_cut) %>% pull(gene_id) %>% unique()
  } else {
    df %>% filter(padj_use < padj_cut, abs(log2FoldChange) >= lfc_cut) %>% pull(gene_id) %>% unique()
  }
}

save_upset_pair <- function(list_input, out_stub, text_scale_labels = 3.2) {

  sizes <- sapply(list_input, length)
  message("UpSet input sizes for ", out_stub, ": ",
          paste(names(sizes), sizes, sep="=", collapse="; "))

  if (all(sizes == 0)) return(invisible(NULL))

  upset_input <- UpSetR::fromList(lapply(list_input, function(x) unique(as.character(x))))

  # show in knitted HTML
  print(UpSetR::upset(
    upset_input,
    order.by = "freq",
    keep.order = TRUE,
    point.size = 3,
    line.size = 1.5,
    mainbar.y.label = "",
    sets.x.label = "",
    text.scale = text_scale_labels
  ))

  # labeled (save)
  grDevices::png(paste0(out_stub, "_labels.png"), width = 2000, height = 1400, res = 200)
  on.exit(grDevices::dev.off(), add = TRUE)

  print(UpSetR::upset(
    upset_input,
    order.by = "freq",
    keep.order = TRUE,
    point.size = 3,
    line.size = 1.5,
    mainbar.y.label = "",
    sets.x.label = "",
    text.scale = text_scale_labels
  ))

  grDevices::dev.off()
  on.exit(NULL, add = FALSE)

  # no-labels (save)
  grDevices::png(paste0(out_stub, "_nolabels.png"), width = 2000, height = 1400, res = 200)
  on.exit(grDevices::dev.off(), add = TRUE)

  print(UpSetR::upset(
    upset_input,
    order.by = "freq",
    keep.order = TRUE,
    point.size = 3,
    line.size = 1.5,
    mainbar.y.label = "",
    sets.x.label = "",
    text.scale = 0
  ))

  grDevices::dev.off()
  on.exit(NULL, add = FALSE)

  invisible(upset_input)
}
```

excel file with DEGs: helpers only
```{r}
res_to_table <- function(res_obj, comparison_name, padj_cut = 0.05, lfc_cut = 1) {
  as.data.frame(res_obj) %>%
    rownames_to_column("gene_id") %>%
    mutate(
      comparison = comparison_name,
      padj_use = if_else(is.na(padj), 1, padj),
      sig_padj = padj_use < padj_cut,
      sig_padj_lfc = padj_use < padj_cut & abs(log2FoldChange) >= lfc_cut
    )
}

```

Export and print plots baseline: controls
```{r}
# 1) Control-only: Harlan Control vs Knoxville Control
df_ctrl <- make_volcano_df(deRes.controls.harlan_vs_knoxville)

save_volcano_pair(
  df_ctrl,
  file.path(outdir, "volcano_Harlan_vs_Knoxville_Control"),
  top_n = top_n_labels, padj_cut = padj_cut, lfc_cut = lfc_cut
)

```

Export and print plots baseline: Deltamethrin

```{r}
# 2) Deltamethrin-only: Harlan Deltamethrin vs Knoxville Deltamethrin
df_delHK <- make_volcano_df(deRes.deltamethrin.harlan_vs_knoxville)

save_volcano_pair(
  df_delHK,
  file.path(outdir, "volcano_Harlan_vs_Knoxville_Deltamethrin"),
  top_n = top_n_labels, padj_cut = padj_cut, lfc_cut = lfc_cut
)

```

Export: Harlan control vs. all treatments

```{r}
# 3) Harlan: Control vs Treatments (UpSet all + 1 volcano per treatment)
upset_harlan_list <- list()
for (trt in treatments_fig) {
  res_obj <- get(paste0("deRes.", trt, ".harlan"))
  upset_harlan_list[[trt]] <- deg_ids_from_res(
    res_obj,
    padj_cut = padj_cut,
    lfc_cut  = lfc_cut   #(padj + |log2FC|>=1)
  )
}
message("Harlan DEG counts (padj<0.05 & |LFC|>=1): ",
        paste(names(upset_harlan_list), sapply(upset_harlan_list, length), collapse = "; "))
save_upset_pair(upset_harlan_list, file.path(outdir, "upSet_Harlan_Control_vs_Treatments"))


for (trt in treatments_fig) {
  dfH <- make_volcano_df(get(paste0("deRes.", trt, ".harlan")))
  save_volcano_pair(
    dfH,
    file.path(outdir, paste0("volcano_Harlan_Control_vs_", trt)),
    top_n = top_n_labels, padj_cut = padj_cut, lfc_cut = lfc_cut
  )
}

```


Export: Knoxville control vs. all treatments 
```{r}
# 4) Knoxville: Control vs Treatments (UpSet all + 1 volcano per treatment)

upset_knox_list <- list()
for (trt in treatments_fig) {
  res_obj <- get(paste0("deRes.", trt, ".knoxville"))
  upset_knox_list[[trt]] <- deg_ids_from_res(
    res_obj,
    padj_cut = padj_cut,
    lfc_cut  = lfc_cut   #(padj + |log2FC|>=1)
  )
}
message("Knoxville DEG counts (padj<0.05 & |LFC|>=1): ",
        paste(names(upset_knox_list), sapply(upset_knox_list, length), collapse = "; "))
save_upset_pair(upset_knox_list, file.path(outdir, "upSet_Knoxville_Control_vs_Treatments"))


for (trt in treatments_fig) {
  dfK <- make_volcano_df(get(paste0("deRes.", trt, ".knoxville")))
  save_volcano_pair(
    dfK,
    file.path(outdir, paste0("volcano_Knoxville_Control_vs_", trt)),
    top_n = top_n_labels, padj_cut = padj_cut, lfc_cut = lfc_cut
  )
}
```

Export .csv files with DEGs
```{r}
# 5) EXPORTS: ALL GENES + DEGs across ALL comparisons
tables_all <- list()

# Control H vs K
tables_all[["Harlan_Control_vs_Knoxville_Control"]] <-
  res_to_table(deRes.controls.harlan_vs_knoxville, "Harlan_Control_vs_Knoxville_Control",
               padj_cut = padj_cut, lfc_cut = lfc_cut)

# Deltamethrin-only H vs K
tables_all[["Harlan_Deltamethrin_vs_Knoxville_Deltamethrin"]] <-
  res_to_table(deRes.deltamethrin.harlan_vs_knoxville, "Harlan_Deltamethrin_vs_Knoxville_Deltamethrin",
               padj_cut = padj_cut, lfc_cut = lfc_cut)

# Harlan Control vs treatments
for (trt in treatments_fig) {
  tables_all[[paste0("Harlan_Control_vs_", trt)]] <-
    res_to_table(get(paste0("deRes.", trt, ".harlan")),
                 paste0("Harlan_Control_vs_", trt),
                 padj_cut = padj_cut, lfc_cut = lfc_cut)
}

# Knoxville Control vs treatments
for (trt in treatments_fig) {
  tables_all[[paste0("Knoxville_Control_vs_", trt)]] <-
    res_to_table(get(paste0("deRes.", trt, ".knoxville")),
                 paste0("Knoxville_Control_vs_", trt),
                 padj_cut = padj_cut, lfc_cut = lfc_cut)
}

all_genes_all_comparisons <- dplyr::bind_rows(tables_all)

write.csv(all_genes_all_comparisons, file.path(outdir, "ALL_genes_all_comparisons.csv"), row.names = FALSE)
openxlsx::write.xlsx(all_genes_all_comparisons, file.path(outdir, "ALL_genes_all_comparisons.xlsx"), rowNames = FALSE)

degs_padj <- all_genes_all_comparisons %>% dplyr::filter(sig_padj)
write.csv(degs_padj, file.path(outdir, "DEGs_padj0.05_all_comparisons.csv"), row.names = FALSE)
openxlsx::write.xlsx(degs_padj, file.path(outdir, "DEGs_padj0.05_all_comparisons.xlsx"), rowNames = FALSE)

degs_padj_lfc <- all_genes_all_comparisons %>% dplyr::filter(sig_padj_lfc)
write.csv(degs_padj_lfc, file.path(outdir, "DEGs_padj0.05_LFC1_all_comparisons.csv"), row.names = FALSE)
openxlsx::write.xlsx(degs_padj_lfc, file.path(outdir, "DEGs_padj0.05_LFC1_all_comparisons.xlsx"), rowNames = FALSE)

message("DONE: Plots + UpSets + exports saved to: ", outdir)
```
```{r}
count_degs <- function(res_obj, padj_cut=0.05, p_cut=0.05, lfc_cut=1) {
  df <- as.data.frame(res_obj) %>%
    tibble::rownames_to_column("gene_id") %>%
    dplyr::mutate(
      padj_use = dplyr::if_else(is.na(padj), 1, padj),
      p_use    = dplyr::if_else(is.na(pvalue), 1, pvalue)
    ) %>%
    dplyr::filter(!is.na(log2FoldChange))

  c(
    padj_only      = sum(df$padj_use < padj_cut),
    padj_lfc       = sum(df$padj_use < padj_cut & abs(df$log2FoldChange) >= lfc_cut),
    pvalue_only    = sum(df$p_use < p_cut),
    pvalue_lfc     = sum(df$p_use < p_cut & abs(df$log2FoldChange) >= lfc_cut)
  )
}

get_counts_for_strain <- function(strain) {
  out <- lapply(treatments_fig, function(trt) {
    res_obj <- get(paste0("deRes.", trt, ".", strain))
    count_degs(res_obj, padj_cut=padj_cut, p_cut=0.05, lfc_cut=lfc_cut)
  })
  out <- do.call(rbind, out)
  rownames(out) <- treatments_fig
  out
}

cat("\nHarlan counts:\n")
print(get_counts_for_strain("harlan"))

cat("\nKnoxville counts:\n")
print(get_counts_for_strain("knoxville"))
```

```{r}
resK_del <- get("deRes.Deltamethrin.knoxville")
dfK_del <- as.data.frame(resK_del) %>%
  tibble::rownames_to_column("gene_id") %>%
  dplyr::mutate(padj_use = dplyr::if_else(is.na(padj), 1, padj)) %>%
  dplyr::arrange(padj_use)

head(dfK_del[, c("gene_id","log2FoldChange","pvalue","padj_use")], 15)
```

curious about acanthoscurrin-gene
```{r}
subset(all_genes_all_comparisons,
       gene_id %in% c("gene-LOC106664987", "gene-LOC106669600")) %>%
  dplyr::select(comparison, log2FoldChange, pvalue, padj_use, sig_padj)
```

```{r}
colnames(all_genes_all_comparisons)
```
```{r}
all_genes_all_comparisons %>%
  dplyr::filter(gene_id %in% c("gene-LOC106664987", "gene-LOC106669600")) %>%
  dplyr::select(gene_id, comparison, log2FoldChange, pvalue, padj_use, sig_padj, sig_padj_lfc) %>%
  dplyr::arrange(gene_id, comparison)
```

