---
title: "bedbug_plots_2026"
author: "Montse Robles"
date: "2026-02-25"
output: html_document
---

```{r setup, include=FALSE}
setwd("/Users/roblesgi/Documents/roblesgi/bedbug/rnaseq/")
outdir <- "Figures_2026"
  library (tidyverse)
  library(ggplot2)
  library(ggrepel)
  library(UpSetR)
  library(openxlsx)
  library(dplyr)
  library(tibble)
  library(grid)
```

We're creating a new folder for the final plots 2026 so we can then edit them on Afinnity and photoshop

```{r}
load("deseq_objects_for_figures.RData")
```

Settings for the comparisons
```{r}
treatments_fig <- c("Carvacrol", "Deltamethrin", "PBO", "Thymol")
padj_cut <- 0.05
lfc_cut  <- 1
top_n_labels <- 10
```

helpers for themes
```{r}
theme_clean <- function() {
  theme_bw() +
    theme(
      panel.grid = element_blank(),
      plot.title = element_blank(),
      legend.position = "none"
    )
}
```

Volcano plot settings (themes)

```{r}
# VOLCANO HELPERS (pvalue on Y, highlight DEGs with padj)
# 1) Build volcano plotting data

make_volcano_df <- function(res_obj) {

  df <- as.data.frame(res_obj) %>%
    rownames_to_column("gene_id")

  # make sure pvalue/padj are numeric
  df <- df %>%
    mutate(
      pvalue = as.numeric(pvalue),
      padj   = as.numeric(padj),
      log2FoldChange = as.numeric(log2FoldChange)
    )

  # use raw p-value on y axis, but avoid Inf
  min_nonzero_p <- suppressWarnings(min(df$pvalue[df$pvalue > 0], na.rm = TRUE))
  if (!is.finite(min_nonzero_p)) min_nonzero_p <- 1e-300

  df <- df %>%
    mutate(
      pvalue_plot = case_when(
        is.na(pvalue)      ~ NA_real_,
        pvalue <= 0        ~ min_nonzero_p,
        TRUE               ~ pvalue
      ),
      neglog10_p = -log10(pvalue_plot),
      padj_plot  = padj
    )

  return(df)
}

#----------------------------------
# 2) Make volcano plot
#----------------------------------
make_volcano_plot <- function(df,
                              padj_cut = 0.05,
                              lfc_cut  = 1,
                              top_n = 8,
                              label_genes = NULL,
                              xlim_use = NULL,
                              point_size_ns = 2.0,
                              point_size_sig = 2.2) {

  # classify points by DEG definition
  df <- df %>%
    mutate(
      sig_class = case_when(
        !is.na(padj_plot) & padj_plot < padj_cut & log2FoldChange >=  lfc_cut ~ "Up",
        !is.na(padj_plot) & padj_plot < padj_cut & log2FoldChange <= -lfc_cut ~ "Down",
        TRUE ~ "NS"
      )
    )

  # labeling logic:
  # if label_genes is given, label only those
  # otherwise label top_n DEGs by padj, then by |LFC|
  if (!is.null(label_genes)) {
    label_df <- df %>%
      filter(gene_id %in% label_genes)
  } else {
    label_df <- df %>%
      filter(
        !is.na(padj_plot),
        padj_plot < padj_cut,
        !is.na(log2FoldChange),
        abs(log2FoldChange) >= lfc_cut
      ) %>%
      arrange(padj_plot, desc(abs(log2FoldChange))) %>%
      slice_head(n = top_n)
  }

  p <- ggplot(df, aes(x = log2FoldChange, y = neglog10_p)) +

    # NS points first
    geom_point(
      data = df %>% filter(sig_class == "NS"),
      color = "grey78",
      alpha = 0.45,
      size = point_size_ns,
      na.rm = TRUE
    ) +

    # Down
    geom_point(
      data = df %>% filter(sig_class == "Down"),
      color = "#3B6EA8",
      alpha = 0.90,
      size = point_size_sig,
      na.rm = TRUE
    ) +

    # Up
    geom_point(
      data = df %>% filter(sig_class == "Up"),
      color = "#C43C39",
      alpha = 0.90,
      size = point_size_sig,
      na.rm = TRUE
    ) +

    # only FC cutoff lines
    geom_vline(
      xintercept = c(-lfc_cut, lfc_cut),
      linetype = "dashed",
      linewidth = 0.5,
      color = "grey35"
    ) +

    # labels
    geom_text_repel(
      data = label_df,
      aes(label = gene_id),
      size = 4.0,
      color = "black",
      box.padding = 0.35,
      point.padding = 0.20,
      segment.color = "grey45",
      segment.size = 0.35,
      min.segment.length = 0,
      max.overlaps = Inf,
      seed = 1,
      na.rm = TRUE
    ) +

    labs(
      x = expression(log[2]~"fold change"),
      y = expression(-log[10]~italic(p)~"value")
    ) +

    theme_classic(base_size = 16) +
    theme(
      axis.title = element_text(size = 18, color = "black"),
      axis.text  = element_text(size = 15, color = "black"),
      axis.line  = element_line(linewidth = 0.8, color = "black"),
      axis.ticks = element_line(linewidth = 0.8, color = "black"),
      axis.ticks.length = unit(0.18, "cm"),
      legend.position = "none"
    )

  if (!is.null(xlim_use)) {
    p <- p + coord_cartesian(xlim = xlim_use)
  }

  return(p)
}

# 3) Save labeled + unlabeled versions
save_volcano_pair <- function(df,
                              outprefix,
                              top_n = 8,
                              padj_cut = 0.05,
                              lfc_cut = 1,
                              label_genes = NULL,
                              xlim_use = NULL,
                              width = 8,
                              height = 6,
                              dpi = 300) {

  # labeled version
  p_labels <- make_volcano_plot(
    df = df,
    padj_cut = padj_cut,
    lfc_cut = lfc_cut,
    top_n = top_n,
    label_genes = label_genes,
    xlim_use = xlim_use
  )

  # unlabeled version
  p_nolabels <- make_volcano_plot(
    df = df,
    padj_cut = padj_cut,
    lfc_cut = lfc_cut,
    top_n = 0,
    label_genes = character(0),
    xlim_use = xlim_use
  )

  ggsave(
    filename = paste0(outprefix, "_labels.png"),
    plot = p_labels,
    width = width,
    height = height,
    dpi = dpi,
    bg = "white"
  )

  ggsave(
    filename = paste0(outprefix, "_labels.pdf"),
    plot = p_labels,
    width = width,
    height = height,
    bg = "white"
  )

  ggsave(
    filename = paste0(outprefix, "_nolabels.png"),
    plot = p_nolabels,
    width = width,
    height = height,
    dpi = dpi,
    bg = "white"
  )

  ggsave(
    filename = paste0(outprefix, "_nolabels.pdf"),
    plot = p_nolabels,
    width = width,
    height = height,
    bg = "white"
  )

  invisible(list(labels = p_labels, nolabels = p_nolabels))
}
```

UpSet helpers
```{r}
# UpSet HELPERS (original-style, reliable file output + render in HTML)

deg_ids_from_res <- function(res_obj, padj_cut = 0.05, lfc_cut = NULL) {
  df <- as.data.frame(res_obj) %>%
    rownames_to_column("gene_id") %>%
    mutate(padj_use = if_else(is.na(padj), 1, padj))

  if (is.null(lfc_cut)) {
    df %>% filter(padj_use < padj_cut) %>% pull(gene_id) %>% unique()
  } else {
    df %>% filter(padj_use < padj_cut, abs(log2FoldChange) >= lfc_cut) %>% pull(gene_id) %>% unique()
  }
}

save_upset_pair <- function(list_input, out_stub, text_scale_labels = 3.2) {

  sizes <- sapply(list_input, length)
  message("UpSet input sizes for ", out_stub, ": ",
          paste(names(sizes), sizes, sep="=", collapse="; "))

  if (all(sizes == 0)) return(invisible(NULL))

  upset_input <- UpSetR::fromList(lapply(list_input, function(x) unique(as.character(x))))

  # show in knitted HTML
  print(UpSetR::upset(
    upset_input,
    order.by = "freq",
    keep.order = TRUE,
    point.size = 3,
    line.size = 1.5,
    mainbar.y.label = "",
    sets.x.label = "",
    text.scale = text_scale_labels
  ))

  # labeled (save)
  grDevices::png(paste0(out_stub, "_labels.png"), width = 2000, height = 1400, res = 200)
  on.exit(grDevices::dev.off(), add = TRUE)

  print(UpSetR::upset(
    upset_input,
    order.by = "freq",
    keep.order = TRUE,
    point.size = 3,
    line.size = 1.5,
    mainbar.y.label = "",
    sets.x.label = "",
    text.scale = text_scale_labels
  ))

  grDevices::dev.off()
  on.exit(NULL, add = FALSE)

  # no-labels (save)
  grDevices::png(paste0(out_stub, "_nolabels.png"), width = 2000, height = 1400, res = 200)
  on.exit(grDevices::dev.off(), add = TRUE)

  print(UpSetR::upset(
    upset_input,
    order.by = "freq",
    keep.order = TRUE,
    point.size = 3,
    line.size = 1.5,
    mainbar.y.label = "",
    sets.x.label = "",
    text.scale = 0
  ))

  grDevices::dev.off()
  on.exit(NULL, add = FALSE)

  invisible(upset_input)
}
```

excel file with DEGs: helpers only
```{r}
res_to_table <- function(res_obj, comparison_name, padj_cut = 0.05, lfc_cut = 1) {
  as.data.frame(res_obj) %>%
    rownames_to_column("gene_id") %>%
    mutate(
      comparison = comparison_name,
      padj_use = if_else(is.na(padj), 1, padj),
      sig_padj = padj_use < padj_cut,
      sig_padj_lfc = padj_use < padj_cut & abs(log2FoldChange) >= lfc_cut
    )
}

```

Export and print plots baseline: controls
```{r}
# 1) Control-only: Harlan Control vs Knoxville Control
df_ctrl <- make_volcano_df(deRes.controls.harlan_vs_knoxville)

save_volcano_pair(
  df = df_ctrl,
  outprefix = file.path(outdir, "volcano_Harlan_vs_Knoxville_Control"),
  top_n = top_n_labels,
  padj_cut = padj_cut,
  lfc_cut = lfc_cut
)
```

Export and print plots baseline: Deltamethrin

```{r}
# genes to label in deltamethrin volcano
labels_deltamethrin <- c(
  "gene-LOC112128126",  # CYP9f2
  "gene-LOC106673503",  # CYP6a18
  "gene-LOC106670429",  # CYP6j1-like
  "gene-LOC106673482",  # CYP6k1-like
  "gene-LOC106673892",  # CYP6d5
  "gene-LOC106665265",  # cuticle
  "gene-LOC106665190",  # cuticle
  "gene-LOC106664847",  # cuticle
  "gene-LOC106663700",  # lysozyme
  "gene-LOC106669244",  # ACP-20-like
  "gene-LOC106670669"   # strong Harlan gene
)
# 2) Deltamethrin-only: Harlan Deltamethrin vs Knoxville Deltamethrin
df_delHK <- make_volcano_df(deRes.deltamethrin.harlan_vs_knoxville)

save_volcano_pair(
  df = df_delHK,
  outprefix = file.path(outdir, "volcano_Harlan_vs_Knoxville_Deltamethrin"),
  padj_cut = padj_cut,
  lfc_cut = lfc_cut,
  label_genes = labels_deltamethrin
)
```

Export: Harlan control vs. all treatments

```{r}
# 3) Harlan: Control vs Treatments (UpSet all + 1 volcano per treatment)
upset_harlan_list <- list()
for (trt in treatments_fig) {
  res_obj <- get(paste0("deRes.", trt, ".harlan"))
  upset_harlan_list[[trt]] <- deg_ids_from_res(
    res_obj,
    padj_cut = padj_cut,
    lfc_cut  = lfc_cut   # DEG = padj < 0.05 & |log2FC| >= 1
  )
}

message(
  "Harlan DEG counts (padj<0.05 & |LFC|>=1): ",
  paste(names(upset_harlan_list),
        sapply(upset_harlan_list, length),
        collapse = "; ")
)

save_upset_pair(
  upset_harlan_list,
  file.path(outdir, "upSet_Harlan_Control_vs_Treatments")
)

for (trt in treatments_fig) {
  dfH <- make_volcano_df(get(paste0("deRes.", trt, ".harlan")))

  save_volcano_pair(
    df = dfH,
    outprefix = file.path(outdir, paste0("volcano_Harlan_Control_vs_", trt)),
    top_n = top_n_labels,
    padj_cut = padj_cut,
    lfc_cut = lfc_cut
  )
}

```


Export: Knoxville control vs. all treatments 
```{r}
# 4) Knoxville: Control vs Treatments (UpSet all + 1 volcano per treatment)

upset_knox_list <- list()
for (trt in treatments_fig) {
  res_obj <- get(paste0("deRes.", trt, ".knoxville"))
  
  upset_knox_list[[trt]] <- deg_ids_from_res(
    res_obj,
    padj_cut = padj_cut,
    lfc_cut  = lfc_cut   # DEG = padj < 0.05 & |log2FC| >= 1
  )
}

message(
  "Knoxville DEG counts (padj<0.05 & |LFC|>=1): ",
  paste(names(upset_knox_list),
        sapply(upset_knox_list, length),
        collapse = "; ")
)

save_upset_pair(
  upset_knox_list,
  file.path(outdir, "upSet_Knoxville_Control_vs_Treatments")
)


for (trt in treatments_fig) {

  dfK <- make_volcano_df(get(paste0("deRes.", trt, ".knoxville")))

  save_volcano_pair(
    df = dfK,
    outprefix = file.path(outdir, paste0("volcano_Knoxville_Control_vs_", trt)),
    top_n = top_n_labels,
    padj_cut = padj_cut,
    lfc_cut = lfc_cut
  )

}
```

Export .csv files with DEGs
```{r}
# 5) EXPORTS: ALL GENES + DEGs across ALL comparisons
tables_all <- list()

# Control H vs K
tables_all[["Harlan_Control_vs_Knoxville_Control"]] <-
  res_to_table(deRes.controls.harlan_vs_knoxville, "Harlan_Control_vs_Knoxville_Control",
               padj_cut = padj_cut, lfc_cut = lfc_cut)

# Deltamethrin-only H vs K
tables_all[["Harlan_Deltamethrin_vs_Knoxville_Deltamethrin"]] <-
  res_to_table(deRes.deltamethrin.harlan_vs_knoxville, "Harlan_Deltamethrin_vs_Knoxville_Deltamethrin",
               padj_cut = padj_cut, lfc_cut = lfc_cut)

# Harlan Control vs treatments
for (trt in treatments_fig) {
  tables_all[[paste0("Harlan_Control_vs_", trt)]] <-
    res_to_table(get(paste0("deRes.", trt, ".harlan")),
                 paste0("Harlan_Control_vs_", trt),
                 padj_cut = padj_cut, lfc_cut = lfc_cut)
}

# Knoxville Control vs treatments
for (trt in treatments_fig) {
  tables_all[[paste0("Knoxville_Control_vs_", trt)]] <-
    res_to_table(get(paste0("deRes.", trt, ".knoxville")),
                 paste0("Knoxville_Control_vs_", trt),
                 padj_cut = padj_cut, lfc_cut = lfc_cut)
}

all_genes_all_comparisons <- dplyr::bind_rows(tables_all)

write.csv(all_genes_all_comparisons, file.path(outdir, "ALL_genes_all_comparisons.csv"), row.names = FALSE)
openxlsx::write.xlsx(all_genes_all_comparisons, file.path(outdir, "ALL_genes_all_comparisons.xlsx"), rowNames = FALSE)

degs_padj <- all_genes_all_comparisons %>% dplyr::filter(sig_padj)
write.csv(degs_padj, file.path(outdir, "DEGs_padj0.05_all_comparisons.csv"), row.names = FALSE)
openxlsx::write.xlsx(degs_padj, file.path(outdir, "DEGs_padj0.05_all_comparisons.xlsx"), rowNames = FALSE)

degs_padj_lfc <- all_genes_all_comparisons %>% dplyr::filter(sig_padj_lfc)
write.csv(degs_padj_lfc, file.path(outdir, "DEGs_padj0.05_LFC1_all_comparisons.csv"), row.names = FALSE)
openxlsx::write.xlsx(degs_padj_lfc, file.path(outdir, "DEGs_padj0.05_LFC1_all_comparisons.xlsx"), rowNames = FALSE)

message("DONE: Plots + UpSets + exports saved to: ", outdir)
```
```{r}
count_degs <- function(res_obj, padj_cut=0.05, p_cut=0.05, lfc_cut=1) {
  df <- as.data.frame(res_obj) %>%
    tibble::rownames_to_column("gene_id") %>%
    dplyr::mutate(
      padj_use = dplyr::if_else(is.na(padj), 1, padj),
      p_use    = dplyr::if_else(is.na(pvalue), 1, pvalue)
    ) %>%
    dplyr::filter(!is.na(log2FoldChange))

  c(
    padj_only      = sum(df$padj_use < padj_cut),
    padj_lfc       = sum(df$padj_use < padj_cut & abs(df$log2FoldChange) >= lfc_cut),
    pvalue_only    = sum(df$p_use < p_cut),
    pvalue_lfc     = sum(df$p_use < p_cut & abs(df$log2FoldChange) >= lfc_cut)
  )
}

get_counts_for_strain <- function(strain) {
  out <- lapply(treatments_fig, function(trt) {
    res_obj <- get(paste0("deRes.", trt, ".", strain))
    count_degs(res_obj, padj_cut=padj_cut, p_cut=0.05, lfc_cut=lfc_cut)
  })
  out <- do.call(rbind, out)
  rownames(out) <- treatments_fig
  out
}

cat("\nHarlan counts:\n")
print(get_counts_for_strain("harlan"))

cat("\nKnoxville counts:\n")
print(get_counts_for_strain("knoxville"))
```

```{r}
resK_del <- get("deRes.Deltamethrin.knoxville")
dfK_del <- as.data.frame(resK_del) %>%
  tibble::rownames_to_column("gene_id") %>%
  dplyr::mutate(padj_use = dplyr::if_else(is.na(padj), 1, padj)) %>%
  dplyr::arrange(padj_use)

head(dfK_del[, c("gene_id","log2FoldChange","pvalue","padj_use")], 15)
```

curious about acanthoscurrin-gene
```{r}
subset(all_genes_all_comparisons,
       gene_id %in% c("gene-LOC106664987", "gene-LOC106669600")) %>%
  dplyr::select(comparison, log2FoldChange, pvalue, padj_use, sig_padj)
```

```{r}
colnames(all_genes_all_comparisons)
```
```{r}
all_genes_all_comparisons %>%
  dplyr::filter(gene_id %in% c("gene-LOC106664987", "gene-LOC106669600")) %>%
  dplyr::select(gene_id, comparison, log2FoldChange, pvalue, padj_use, sig_padj, sig_padj_lfc) %>%
  dplyr::arrange(gene_id, comparison)
```

